---
title: "Fun with music: Exploring how words in pop songs have changed over time"
output: github_document

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#### Sources

* Pop songs and their ranking: Pop songs for each year were gathered from Billboard's top 100 rankings found here: https://www.billboard.com/charts/hot-100
For each song, the date when the song peaked and the peak ranking on Billboard was gathered from each song's page, for example: https://www.billboard.com/charts/hot-100?rank=1

* Lyrics: All lyrics were gathered from AZLyrics.com

#### Getting the data

The file "01 Scrapping Songs" scrapes Billboard for the top 100 songs in each year from 2009 to 2019. From Billboard I can get the peak rank for the song, it's peak date and highest rank position in the year and supporting artists.
Then, each song + artist combo is searched on AZLyrics to create the lyrics dataset. Some songs were associated with the supporting artists on AZLyrics only so I re-ran the script to get more matches.

**Total songs and artists by year gathered from Billboard:** 


```{r, echo = FALSE, message=FALSE}

library(tidyverse, quietly = T)
library(knitr)
lyricsdata <- readRDS("AZ_data.rds")
full_list <- readRDS("Billboard_data.rds") %>% 
  mutate(peak_date= lubridate::ymd(gsub("\".*", "", peak_date)))
all_words_complete <- readRDS("Lyrics_data.rds") %>% 
  ungroup() %>% 
  mutate_at(vars("rank", "year", "ID"), as.numeric)

x = full_list %>% 
  mutate(lyrics_found= ifelse(ID %in% lyricsdata$ID, 1, 0),
         peak_date_found = ifelse(!is.na(peak_date), 1, 0)) %>% 
  group_by(year) %>% 
  summarize(`Total Songs`= n(), 
            `Total Songs`= n_distinct(songtitle, artist1, artist2, artist3),
            `Distinct Artists`= n_distinct(c(artist1, artist2, artist3)),
            `Songs With Lyrics`= sum(lyrics_found),
            `Songs with Peak Dates`= sum(peak_date_found)) %>% 
  rename(Year= year)

kable(x) 
```

**Notes:** 

* There is an extra song in 2012. This is because Christina Perri's song "A Thousand Years" comes in two parts and is being treated as one entry. For now I will keep these songs separate despite having the same ranking
* For some reason 2011 and 2016 have missing songs. In 2011, number 7 is missing and in 2016, number 87 is missing
* Peak date information was not available for all songs but ranges from 70% to 95% depending on year

#### Looking at top artists over time

The top 10 most popular artists across the data:

```{r, echo = FALSE, message=FALSE}

artists= full_list %>% 
  select(artist1) %>% 
  rbind(full_list %>% 
          select(artist2) %>% rename(artist1=artist2)) %>% 
  rbind(full_list %>% 
          select(artist3) %>% rename(artist1= artist3))

knitr::kable(artists %>% 
  count(artist1) %>% 
  filter(!artist1=="") %>% 
  arrange(desc(n)) %>% head(10) %>% 
  mutate(artist1= stringr::str_to_title(artist1),
         artist1= case_when(
           artist1== "Nickiminaj"~"Nicki Minaj",
           artist1=="Taylorswift" ~"Taylor Swift",
           artist1=="Lilwayne"~ "Lil Wayne",
           artist1=="Maroon5"~"Maroon 5",
           artist1=="Arianagrande"~ "Ariana Grande",
           artist1=="Brunomars"~"Bruno Mars",
           artist1=="Justinbieber"~  "Justin Bieber",
           artist1=="Chrisbrown"~"Chris Brown",
           artist1=="Justintimberlake"~"Justin Timberlake",
           TRUE~as.character(artist1)
         )
  ) %>% 
  rename(Artist=artist1, `Total Top Songs`=n)
)

```


Drake is consinderably more popular than the next artist in the list with over 50 songs making the top 100 list.

How does the success of top artists track over time?

```{r, echo = FALSE, message=FALSE}
library(ggplot2)
library(ggthemes)
library(RColorBrewer)

nb.cols <- 10
artists_2= artists %>%
  count(artist1) %>% 
  filter(!artist1=="") %>% 
  arrange(desc(n)) %>% head(10)


full_list %>% 
  gather(artist_number, artist_comp, artist1:artist3) %>% 
  mutate(top10_artist= ifelse(artist_comp %in% artists_2$artist1, 1, 0)) %>% 
  filter(top10_artist==1) %>% 
  count(year, artist_comp) %>% 
  rename(Artist=artist_comp, Year=year) %>% 
  mutate(Artist= stringr::str_to_title(Artist),
         Artist= case_when(
           Artist== "Nickiminaj"~"Nicki Minaj",
           Artist=="Taylorswift" ~"Taylor Swift",
           Artist=="Lilwayne"~ "Lil Wayne",
           Artist=="Maroon5"~"Maroon 5",
           Artist=="Arianagrande"~ "Ariana Grande",
           Artist=="Brunomars"~"Bruno Mars",
           Artist=="Justinbieber"~  "Justin Bieber",
           Artist=="Chrisbrown"~"Chris Brown",
           Artist=="Justintimberlake"~"Justin Timberlake",
           TRUE~as.character(Artist)
         )
  ) %>% 
  ggplot(., aes(x=Year, y=n, fill=Artist)) + geom_bar(stat="identity", position= "stack") + scale_fill_tableau(palette = "Tableau 10") +theme_gdocs() +ylab("Total Songs in Top 100 List") + 
  labs(title= "Number of Songs from The Top 10 Artist by Year") +
  theme(plot.title = element_text(hjust = 0.5), 
    legend.position = "right")+labs(size = 0.5) + 
  theme(plot.title = element_text(size = 14, face= "bold"))


```


#### Looking at the lyrics of songs by year

The lyrics data is pretty messy, it has a lot of "words" like "oh oh oh...". I have tried to filter these out using stop words from tidytext as well as other general cleaning (removing punctuation, spaces etc). 

To start, I want to see whether words fall into "positive" or "negative" categories. I'm using "get_sentiment" from tidytext and Bing Liu and Co's dataset. This dataset classifies words into "positive" or "negative" categories. How does this change over time?

```{r,echo = FALSE, message=FALSE}

lyrics_sum_year = all_words_complete %>% 
  ungroup() %>% 
  filter(stop_word==0) %>% 
  count(sentiment, year) %>%
  mutate(sentiment = case_when(
    sentiment=="negative" ~ "Negative",
    sentiment == "positive"~ "Positive",
     is.na(sentiment) ~ "Neutral/Unlabeled",
    T ~ sentiment)) 

ggplot(lyrics_sum_year) +
 aes(x = year, fill = sentiment, weight= n) +
 geom_bar() +
scale_fill_brewer(palette = "Set1", ) +
 labs(x = "Year", y = "Count of Lyrics", title = "Sentiment of words by year, using Bing Liu and Co's dataset", fill = "Sentiment") +
 ggthemes::theme_gdocs()+
theme(plot.title = element_text(hjust = 0.5), 
    legend.position = "right")+labs(size = 0.5) + 
  theme(plot.title = element_text(size = 14, face= "bold"))



```

I see a lot of words that are marked NA. I'm not sure which ones are NA because they're neutral and which ones are NA because they're not really words like "oooh".
*Problem* Need to solve for this- maybe I can remove words that are non-standard?

```{r,echo = FALSE, message=FALSE}




kable(all_words_complete %>% 
  ungroup() %>% 
  filter(stop_word==0) %>% 
  count(sentiment, year) %>%
  mutate(sentiment = case_when(
    sentiment=="negative" ~ "Negative",
    sentiment == "positive"~ "Positive",
     is.na(sentiment) ~ "Neutral/Unlabeled",
    T ~ sentiment))  %>%
  group_by(year) %>% 
  mutate(perc = paste0(round(100*(n/sum(n)),2), "%")) %>% 
  select(-n) %>% 
    rename(Year=year) %>% 
  spread(sentiment, perc) %>% 
  select(1,2,4,3) %>% 
    left_join(x %>% select(Year, `Songs With Lyrics`), by=c("Year"))
  
  )


```

Taking a look at the most common words and their sentiments: 
```{r,echo = FALSE, message=FALSE}
all_words_complete %>% 
  ungroup() %>% 
  count(lyrics, sentiment) %>% 
  arrange(desc(n)) %>% 
  head(10) %>% 
  rename(Words=lyrics, Sentiment=sentiment, `Total Instances`= n) %>% 
  knitr::kable()

```

Looking at top 10 positive, negative and neutral words: 
```{r,echo = FALSE, message=FALSE}

lyrics_sum = all_words_complete %>% 
  ungroup() %>% 
  filter(stop_word==0) %>% 
  count(lyrics, sentiment) %>%
  mutate(sentiment = case_when(
    sentiment=="negative" ~ "Negative",
    sentiment == "positive"~ "Positive",
    is.na(sentiment) ~ "Neutral/Unlabeled",
    T ~ sentiment)) %>% 
  group_by(sentiment)  %>% 
  top_n(10) %>% 
  arrange( desc(n)) 

ggplot(lyrics_sum) +
 aes(x = reorder(lyrics, n), fill = sentiment, weight = n) +
 geom_bar() +
 scale_fill_brewer(palette = "Set1", ) +
 labs(y = "Count", title = "Most Common Positive, Negative and Neutral words", fill = "Sentiment", x = "Lyrics") +
 coord_flip() +
 ggthemes::theme_gdocs() +
 theme(legend.position = "top") +
 facet_wrap(vars(sentiment), scales = "free_y") + theme(plot.title = element_text(hjust = 0.5), 
    legend.position = "right") +labs(size = 0.5) + 
  theme(plot.title = element_text(size = 14, face= "bold"))
```

Unsurprisingly, most words are just filler words.

Looking at only words marked "positive" and "negative" over time:
```{r,echo = FALSE, message=FALSE}

lyrics_sum_year %>%
 filter(sentiment!= "Neutral/Unlabeled") %>%
 mutate(n= ifelse(sentiment=="Negative", n*-1, n)) %>% 
 filter(!is.na(n)) %>%
 ggplot() +
 aes(x = year, fill = sentiment, weight = n) +
 geom_bar() +
 scale_fill_brewer(palette = "Set1") +
 labs(x = "Year", y = "Count", title = "Total Positive and Negative Words by Year", fill = "Sentiment") +
 coord_flip() +
 ggthemes::theme_gdocs()+
 theme(plot.title = element_text(hjust = 0.5), 
    legend.position = "right") +labs(size = 0.5)+ 
  theme(plot.title = element_text(size = 14, face= "bold"))



```

Was the sentiment of pop songs more positive or negative by year?
```{r,echo = FALSE, message=FALSE}

lyrics_sum_year %>%
 filter(sentiment!= "Neutral/Unlabeled") %>%
  group_by(year) %>% 
 mutate(perc=100*(n/sum(n))) %>% 
# mutate(perc= ifelse(sentiment=="Negative", perc*-1, perc)) %>% 
 ggplot() +
 aes(x = year, fill = sentiment, weight = perc) +
 geom_bar() +
 scale_fill_brewer(palette = "Set1") +
 labs(x = "Year", y = "Percentage of Words with Sentiment", title = "Percentage of Positive and Negative \nWords out of all Words with Sentiment by Year", fill = "Sentiment") +
 coord_flip() +
 ggthemes::theme_gdocs()+
 theme(plot.title = element_text(hjust = 0.5), 
    legend.position = "right") +labs(size = 0.5) + 
  theme(plot.title = element_text(size = 14, face= "bold"))


```

However, this is an oversimplified approach, what if the lyric is "bad love", the sentiment should be negative but will be classified as both negative (bad) and positive (love). Maybe getting a ratio of positive to negative words for each song will be a better assessment.

```{r,echo = FALSE, message=FALSE, warning= FALSE}

ratios = all_words_complete %>% 
  ungroup() %>% 
  left_join(full_list %>%  select(ID, peak_date, peak_position), by="ID") %>% 
  count(peak_date, ID, sentiment, rank ,peak_position) %>% 
  mutate(sentiment = case_when(
    sentiment=="negative" ~ "Negative",
    sentiment == "positive"~ "Positive",
    is.na(sentiment) ~ "Neutral/Unlabeled",
    T ~ sentiment)) %>%  
  spread(sentiment, n) %>% 
  mutate(POSRatio = Positive/Negative,
         ID= as.numeric(ID),
         peak_date= as.numeric(peak_date),
         rank= as.numeric(rank),
         peak_position= as.numeric(peak_position))

ratios %>%
 filter(POSRatio >= 0L & POSRatio <= 23L | is.na(POSRatio)) %>%
 ggplot() +
 aes(x = peak_date, y = POSRatio, colour = rank) +
 geom_point(size = 1L) +
 geom_smooth(span = 0.69) +
 scale_color_distiller(palette = "OrRd") +
 labs(x = "Peak Date in Charts", y = "Positive to Negative Ratio", title = "Positive to Negative words ratio by song", color = "Position in Top 100") +
 ggthemes::theme_gdocs()+
 theme(plot.title = element_text(hjust = 0.5), 
    legend.position = "right") +labs(size = 0.5) + 
  theme(plot.title = element_text(size = 14, face= "bold"))

```


